---
title: "ESM 217 Lab 3: Modeling environment-health relationships"
author: "Maxwell Pepperdine"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
    code-summary: "Show the code"
    theme: flatly
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(tidyverse)
library(lubridate)
library(doBy)
library(mosaic)
library(car)
library(sf)
library(grDevices)
library(sjPlot)
library(here)
```

## Load data

```{r, echo = TRUE, output = FALSE}
#read in dengue case data:
Den <- read.csv(here("data/Dengue_Confirmed_Peru.csv")) # change file path for your context
head(Den)
lapply(Den, class)

# read in MDD temperature data from GEE:
MDD_temp <- read.csv(here("data/MDD_Temperature.csv")) # change file path for your context
head(MDD_temp)

# read in MDD precipitation data from GEE:
MDD_precip <- read.csv(here("data/MDD_Precipitation.csv"))
```

## Data wrangling

#### Dengue data

```{r, echo = TRUE, output = FALSE}
#make Date a date object
Den$Date <- as.Date(Den$Week_Yr)

#floor dates for later merging with environment data:
head(Den)
Den$Date_floored <- floor_date(Den$Date, "week")
head(Den)

Den_plot <- ggplot(Den, aes(x=Date_floored, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")
Den_plot
```

#### MDD temp data

```{r, echo = TRUE, output = FALSE}
#convert system.index to date:
MDD_temp$Date <- substr(MDD_temp$system.index, 1, 8)
head(MDD_temp)
MDD_temp$Date <- gsub('(.{4})(.{2})(.*)', '\\1-\\2-\\3', MDD_temp$Date)
head(MDD_temp)
MDD_temp <- MDD_temp[,c(2:4)]
MDD_temp$Date <- as.Date(MDD_temp$Date)
head(MDD_temp)

# floor dates to week to summarize:
MDD_temp$Date_floored <- floor_date(MDD_temp$Date, "week")
head(MDD_temp)
length(unique(MDD_temp$Date))
length(unique(MDD_temp$Date_floored))

MDD_temp_weekly <- summaryBy(mean ~ IDDIST + Date_floored, data=MDD_temp, FUN=mean)
head(MDD_temp_weekly)
colnames(MDD_temp_weekly)[1] <- "Dist_Code"
colnames(MDD_temp_weekly)[3] <- "Mean_Temp_C"
```

#### Merge the datasets

```{r, echo = TRUE, output = FALSE}
### merge dengue and temperature data by floored dates:
Dengue_Temp <- merge(Den, MDD_temp_weekly, 
                     by=c("Dist_Code", "Date_floored"), 
                     all.x=F, all.y=F)
head(Dengue_Temp)
unique(Dengue_Temp$Year)
```

#### MDD precip data

```{r, echo = TRUE, output = FALSE}
# convert system index to date:
MDD_precip$Date <- substr(MDD_precip$system.index, 1, 8)
head(MDD_precip)
MDD_precip$Date <- gsub('(.{4})(.{2})(.*)', '\\1-\\2-\\3', MDD_precip$Date)
head(MDD_precip)
MDD_precip <- MDD_precip[,c(2:4)]
MDD_precip$Date <- as.Date(MDD_precip$Date)
head(MDD_precip)

# floor dates to week to summarize:
MDD_precip$Date_floored <- floor_date(MDD_precip$Date, "week")
head(MDD_precip)
length(unique(MDD_precip$Date))
length(unique(MDD_precip$Date_floored))

MDD_precip_weekly <- summaryBy(mean ~ IDDIST + Date_floored, data=MDD_precip, FUN=mean)
head(MDD_precip_weekly)
colnames(MDD_precip_weekly)[1] <- "Dist_Code"
colnames(MDD_precip_weekly)[3] <- "Mean_Precip_m"
```

#### Merge the precip dataset to the dengue-temperature dataset

```{r, echo = TRUE, output = FALSE}
### merge dengue-temperature and precipitation data by floored dates:
Dengue_Temp_Precip <- merge(Dengue_Temp, MDD_precip_weekly, 
                            by=c("Dist_Code", "Date_floored"), 
                            all.x=F, all.y=F)
head(Dengue_Temp_Precip)
```

## Data analysis

#### Only Dengue cases & temperature

```{r, echo=TRUE, output=FALSE}
# standardize variables:
Dengue_Temp$Mean_Temp_C_std <- zscore(Dengue_Temp$Mean_Temp_C, na.rm=T)
hist(Dengue_Temp$Mean_Temp_C, n=50)
hist(Dengue_Temp$Mean_Temp_C_std, n=50)
Dengue_Temp$Cases_std <- zscore(Dengue_Temp$Cases, na.rm=T)
hist(Dengue_Temp$Cases, n=50)
hist(Dengue_Temp$Cases_std, n=50)
Dengue_Temp$log.Mean_Temp_C <- log(Dengue_Temp$Mean_Temp_C)
hist(Dengue_Temp$Mean_Temp_C, n=50)
hist(Dengue_Temp$log.Mean_Temp_C, n=50)
Dengue_Temp$log.Cases <- log(Dengue_Temp$Cases + 1)
hist(Dengue_Temp$Cases, n=50)
hist(Dengue_Temp$log.Cases, n=50)
```

```{r}
#model dengue cases by temperature:
#head(Dengue_Temp)

Dengue_lm <- lm(
  log.Cases ~
    log.Mean_Temp_C
  + I(log.Mean_Temp_C^2)
  + factor(Year)
  + factor(Dist_Code)
  ,
  data=Dengue_Temp,
  na.action=na.exclude
)
summary(Dengue_lm)
```

```{r}
#| label: fig-dengue-lm
#| fig-cap: "Regression diagnostics for the dengue temperature model."

crPlots(Dengue_lm)
```

```{r}
#| label: fig-dengue-lm-pred
#| fig-cap: "Predicted values from the dengue temperature model."

plot_model(Dengue_lm, type = "pred", 
           terms = "log.Mean_Temp_C", pred.type = "fe")
```

```{r}
#| label: fig-dengue-lm-pred2
#| fig-cap: "Predicted model values for which temperature dengue peaks at (~24.6 C)."

#find at what temperature dengue peaks:
vertex <- -26.75177/(2*-4.17434)
vertex
plot(Dengue_Temp$Mean_Temp_C, Dengue_Temp$log.Mean_Temp_C)
abline(h=3.2)
abline(v=24.6)
#looks like ~25 C, close to predicted thermal optimum for dengue transmitted by Aedes albopictus (25.4-27.6 C) 
```

```{r, echo=TRUE, output=FALSE}
#exploratory plots
Den_plot2 <- ggplot(Dengue_Temp, aes(x=Mean_Temp_C, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")
Den_plot2

Den_plot3 <- ggplot(Dengue_Temp, aes(x=Date_floored, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")
Den_plot3

Den_plot4 <- ggplot(Dengue_Temp, aes(x=Date_floored, y=Mean_Temp_C, col=as.factor(District))) +
  geom_line() #+ theme(legend.position="none")
Den_plot4
```

#### Adding precipitation to the model

```{r, echo=TRUE, output=FALSE}
# standardize variables for temp & cases:
Dengue_Temp_Precip$Mean_Temp_C_std <- zscore(Dengue_Temp_Precip$Mean_Temp_C, na.rm=T)
hist(Dengue_Temp_Precip$Mean_Temp_C, n=50)
hist(Dengue_Temp_Precip$Mean_Temp_C_std, n=50)
Dengue_Temp_Precip$Cases_std <- zscore(Dengue_Temp_Precip$Cases, na.rm=T)
hist(Dengue_Temp_Precip$Cases, n=50)
hist(Dengue_Temp_Precip$Cases_std, n=50)
Dengue_Temp_Precip$log.Mean_Temp_C <- log(Dengue_Temp_Precip$Mean_Temp_C)
hist(Dengue_Temp_Precip$Mean_Temp_C, n=50)
hist(Dengue_Temp_Precip$log.Mean_Temp_C, n=50)
Dengue_Temp_Precip$log.Cases <- log(Dengue_Temp_Precip$Cases + 1)
hist(Dengue_Temp_Precip$Cases, n=50)
hist(Dengue_Temp_Precip$log.Cases, n=50)

# standardize variables for precip:
Dengue_Temp_Precip$Mean_Precip_m_std <- zscore(Dengue_Temp_Precip$Mean_Precip_m, na.rm=T)
hist(Dengue_Temp_Precip$Mean_Precip_m, n=50)
hist(Dengue_Temp_Precip$Mean_Precip_m_std, n=50)
Dengue_Temp_Precip$log.Mean_Precip_m <- log(Dengue_Temp_Precip$Mean_Precip_m + 1)
hist(Dengue_Temp_Precip$Mean_Precip_m, n=50)
hist(Dengue_Temp_Precip$log.Mean_Precip_m, n=50)
```

```{r}
#model dengue cases by temperature and precipitation:
#head(Dengue_Temp_Precip)

Dengue_lm2 <- lm(
  log.Cases ~
    log.Mean_Temp_C
  + I(log.Mean_Temp_C^2)
  + log.Mean_Precip_m
  + I(log.Mean_Precip_m^2)
  + factor(Year)
  + factor(Dist_Code)
  ,
  data=Dengue_Temp_Precip,
  na.action=na.exclude
)

summary(Dengue_lm2)
```

```{r}
#| label: fig-dengue-lm2
#| fig-cap: "Regression diagnostics for the dengue temperature-precipitation model."

crPlots(Dengue_lm2)
```

```{r}
#| label: fig-dengue-lm2-pred
#| fig-cap: "Predicted temperature values from the dengue temperature-precipitation model."

plot_model(Dengue_lm2, type = "pred", 
           terms = "log.Mean_Temp_C", 
           pred.type = "fe")
```

```{r}
#| label: fig-dengue-lm2-pred2
#| fig-cap: "Predicted precipitaton values from the dengue temperature-precipitation model."

plot_model(Dengue_lm2, type = "pred", 
           terms = "log.Mean_Precip_m", 
           pred.type = "fe")
```

```{r, echo=TRUE, output=FALSE}
#exploratory plots
Den_plot5 <- ggplot(Dengue_Temp_Precip, aes(x=Mean_Temp_C, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")
Den_plot5

Den_plot6 <- ggplot(Dengue_Temp_Precip, aes(x=Mean_Precip_m, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")

Den_plot7 <- ggplot(Dengue_Temp_Precip, aes(x=Date_floored, y=Cases, col=as.factor(District))) +
  geom_point() + theme(legend.position="none")
```

## Takeaways

#### Temp only model:

Looking at the regression results when running a model with only temperature, a one unit increase in `log.Mean_Temp_C` increases the dependent variables, `log.Cases`, ~by 26.75 units, holding all else constant. A one unit increase in `log.Mean_Temp_C^2` decreases the dependent variable, `log.Cases`, by ~4.17 units, holding all else constant. This suggests that the effect of temperature on dengue cases is nonlinear with the following traits:

- At lower temperatures, an increase in temperature has a positive effect on dengue cases.
- At higher temperatures, the effect starts to decrease, and an increase in temperature has either a smaller effect or negative effect on dengue cases.

#### Temp & precipitation model:
  
After repeating this analysis with precipitation added to the model, we see that both temperature variables (`log.Mean_Temp_C` & `log.Mean_Temp_C^2`) no longer have a significant effect on dengue cases. Precipitation, however, follows a similar trend to temperature in the temperature only model where: 

- a one unit increase in `log.Mean_Precip_m` increases the dependent variable, `log.Cases`, by ~35.84 units, holding all else constant; and
- a one unit increase in `log.Mean_Precip_m^2` decreases the dependent variable, `log.Cases`, by ~541.3 units, holding all else constant.

This suggests that the effect of precipitation on dengue cases is also nonlinear with the following traits:

- At lower precipitation levels, an increase in precipitation has a positive effect on dengue cases.
- Beyond a certain precipitation level, the effect starts to decrease, and an increase in precipitation has either a smaller effect or negative effect on dengue cases.













